On Optimality of Finite Element Integration

### 2015-09-29 TOMS_Paper/assembly.tex
Trying to rewrite the part on zero block elimination

\subsection{Independence from Form Compilers}
\label{sec:zeros}
COFFEE expects generic abstract syntax trees, so clearly any form compiler can be adapted to produce this kind of input. This is what we have done with the FEniCS Form Compiler, suitably modified to produce abstract syntax trees rather than strings. However, form compilers that translate quadrature-based expressions into code could decide, for simplicity, to implement non-scalar function spaces in trivial yet sub-optimal ways. For example ...by producing a trivial loop nest that iterates over basis function tables filled with blocks of zeros

Another viewpoint is: COFFEE takes care of the optimization process, ...

Consider a set of tabulated basis functions with quadrature points along rows and functions along columns. For example, \texttt{A[i,j]} provides the value of the \texttt{j-th} basis function at quadrature point \texttt{i}. Unless using a smart form compiler (which we want to avoid), there are circumstances in which the tables are sparse. Zero-valued columns arise when taking derivatives on a reference element or when employing vector-valued elements. Zero-valued rows can result from using non-standard functions spaces, like Raviart-Thomas. Zero-valued blocks can appear in pre-evaluated temporaries. Our objective is a transformation that avoids useless iteration over zeros while preserving the effectiveness of the other low-level optimizations, especially vectorization.

In~\citeN{quadrature1}, a technique to avoid iteration over zero-valued columns based on the use of indirection arrays (e.g. \texttt{A[B[i]]}, in which \texttt{A} is a tabulated basis function and \texttt{B} a map from loop iterations to non-zero columns in A) was proposed. Our approach, which will be compared to this pioneering work, aims to free the generated code from such indirection arrays. This is because we want to avoid non-contiguous memory loads and stores, which can nullify the benefits of vectorization.

@@

% say that the reasons to have zeros is to keep the design of the form compiler simple, which just ``puts zeros'' and ``outline the assembly expression''.
% rationale: contiguous in the innermost dimension are aggregated. contiguous in all other dimensions are not, unless complete overlap.
% tables shrunk and merged

>>>>

\subsection{Independence from Form Compilers}
\label{sec:zeros}
One of the key pillars of COFFEE is independence from form compilers. Being capable of handling generic abstract syntax trees, COFFEE can be leveraged by any suitably adapted form compiler. For example, in Firedrake a modified version of the FEniCS Form Compiler producing abstract syntax trees instead of strings is used. COFFEE itself provides an interface for building such abstract syntax trees. In particular, COFFEE aims to decouple the form manipulation (i.e., the ``math level'') from code optimization (flops executed, low level optimization such as vectorization). Another viewpoint is: developers of form compilers should not worry about expression optimization in any way, since this is handled at a lower level.

To achieve this goal, COFFEE needs a strategy to minimize the execution of potentially ``useless'' flops due to simplistic implementations of non scalar-valued function spaces. For example, the most trivial quadrature mode in the FEniCS Form Compiler implements vector-valued function spaces by populating tabulated basis functions with blocks of zero-valued columns. For the local element matrix/vector evaluation, these blocks are iterated over in a single, rectangular iteration space. This preserves the semantics of the computation while keeping the implementation complexity extremely low, although it also introduces unnecessary operations (e.g., zero-valued summands due to multiplications by zero). COFFEE achieves full independence from form compilers (and so relives the implementation burden) by providing a transformation to avoid the execution of such useless flops. A challenge for this transformation is restructuring the iteration spaces while preserving the effectiveness of low level optimizations, especially (auto-)vectorization.

Consider a set of tabulated basis functions with quadrature points along rows and functions along columns. For example, \texttt{A[i,j]} provides the value of the \texttt{j-th} basis function at quadrature point \texttt{i}. In~\citeN{quadrature1}, a technique to avoid iteration over zero-valued columns based on the use of indirection arrays (e.g. \texttt{A[B[i]]}, in which \texttt{A} is a tabulated basis function and \texttt{B} a map from loop iterations to non-zero columns in A) was proposed. Our approach aims to avoid such indirection arrays in the generated code. This is because we want to avoid non-contiguous memory loads and stores, which can nullify the benefits of vectorization.

@@
